Scaling rates of true polar wander in convecting planets and moons

### 2016-06-07 tpw_rate.tex
Small edits, and fix line numbering.

@@

\begin{figure*}
\centering
\includegraphics[width=0.9\textwidth]{figures/reference_frames.pdf}
\caption{Relevant vectors and angles for the TPW analysis. The axes $\mathbf{e}_1$, $\mathbf{e}_2$, and $\mathbf{e}_3$ represent the principal axes of the convective moment of inertia $\mathbf{E}$, with associated eigenvalues $\lambda_1$, $\lambda_2$, and $\lambda_3$, respectively. Since the choice of geographic axes is arbitrary, we assume that at this instant the $\mathbf{E}$-frame and the geographic frames are colocated (though at future times they will not be). The angle $\theta$ represents the mismatch between the rotation axis $\mitbf{\omega}$ and the $\mathbf{e}_3$-axis. For illustration, we assume the longitude $\psi$ of the rotation axis is zero so $\mitbf{\omega}$ lies in the $\mathbf{e_1}$-$\mathbf{e_3}$ plane. True polar wander moves the rotation axis towards the $\mathbf{e}_3$-axis. We denote this motion by $\dot{\mitbf{\omega}}$ in the geographic frame. However, time-dependent convection in the mantle may cause a relative rotation between the geographic frame and the $\mathbf{E}$-frame. This relative rotation can be represented as a rotation around the axis $\mitbf{\Psi}$, which is defined by colatitude $\beta$ and longitude $\gamma$. The rotation around the $\mitbf{\Psi}$ axis contributes to the motion of $\mitbf{\omega}$ as seen in the $\mathbf{E}$-frame.  Further discussion of the reference frames and the relative rotation vector $\mitbf{\Psi}$ can be found in Sections~\ref{sec:reference_frames} and \ref{sec:theta}}
\label{fig:reference_frames}
\end{figure*}

@@

It is useful to define a nondimensional eigenvalue difference (or ``eigengap'') $\Lambda_{ij} \equiv (\lambda_i - \lambda_j)/I_0$.  
This quantity represents the size of fluctuations in the convective moment of inertia compared to the reference value.
The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is 
proportional to the ratio of rotational to gravitational forces (\citep{munk1960rotation}:
\begin{equation}
(C-A) \sim I_0 m.
\end{equation}

@@

Convection drives a mismatch $\theta$ between the current rotation axis and the principal axis of the convective moment. 
This parameter is crucial for the rate of TPW, as described in is in Equation~\eqref{eq:scaled_rotation}.  
Much of the debate around the existence and magnitude of TPW on Earth comes down to the question of how big $\theta$ can be \citep{kirschvink1997evidence, steinberger1997changes}.

Two processes control the evolution of $\theta$. Its growth occurs through perturbations in the convective moment and 
its decay occurs by relaxation of the pole towards the maximum moment of inertia.
We can explore these two effects by converting Equation~\eqref{eq:liouville_ode}
from the body-fixed frame to the $\mathbf{E}$-frame (recall that $\mitbf{\Omega}$ defines
the rotation of the body-fixed frame relative to inertial space in Equation~\eqref{eq:euler}).
The $\mathbf{E}$-frame rotates slowly with respect to the geographic frame, 
which we describe by the rotation vector $\mathbf{\Psi}$. 
(To be explicit in our discussion we let $\mitbf{\Psi}$ define the rotation of the $\mathbf{E}$-frame in the geographic frame, 
although this choice is arbitrary as long as we are consistent). The time derivatives of $\mitbf{\omega}$ in the two frames are related by 
\begin{equation}
\dot{\mitbf{\omega}} = \dot{\mitbf{\omega}}^\prime + \mathbf{\Psi}\times \mitbf{\omega}^\prime.
\end{equation}
where primes are used to define quantities in the $\mathbf{E}$-frame. Rearranging gives 
\begin{equation}
\dot{\mitbf{\omega}}^{\prime} = \dot{\mitbf{\omega}} - \mathbf{\Psi}\times \mitbf{\omega}^\prime.
\end{equation}
We now substitute for $\dot{\mitbf{\omega}}$ from Equation~\eqref{eq:liouville_ode}, 
making the additional assumption that the geographic and $\mathbf{E}$-frames are momentarily aligned. 
This assumption is not a serious restriction because our goal is simply to relate the time derivatives of $\mitbf{\omega}$ in the two frames, 
rather than track the evolution of $\mitbf{\omega}$ over time. The expression for $\dot{\mitbf{\omega}}$ becomes
\begin{equation}
 \dot{\mitbf{\omega}}^\prime  = \frac{1}{(C-A)T_1} \left[ \mathbf{E}^\prime \cdot \mitbf{\omega}^\prime - \left( \mitbf{\omega}^\prime \cdot \mathbf{E}^\prime \cdot \mitbf{\omega}^\prime  \right) \mitbf{\omega}^\prime \right] - \mathbf{\Psi}\times \mitbf{\omega}^\prime.
\label{eq:liouville_ode_new_frame}
\end{equation}
The advantage of expressing the rate of TPW in the $\mathbf{E}$-frame is that the entire equation
may be written in terms of the principal moments $\lambda_i$ and the angles from the principal axes.
As in Equation~\eqref{eq:simple_milankovitch}, we make the simplifying assumption that 
$\mitbf{\omega}^{\prime}$ lies in $\mathbf{e}_1^{\prime}-\mathbf{e}_3^{\prime} $ plane (as shown in Figure~\ref{fig:reference_frames}). 
In this case the orientation of $\mitbf{\omega}$ is defined by colatitude $\theta$ and longitude $\phi=0$. 
The colatitude $\theta$ is precisely the angular misalignment that specifies the rate of TPW in Equation~\eqref{eq:simple_milankovitch}.
We furthermore specify the orientation of $\Psi$
using colatitude $\beta$ and longitude $\gamma$ (see Figure~\ref{fig:reference_frames}).

The time derivative of the unit vector $\mitbf{\omega}^{\prime}$ defines the changing orientation of the rotation axis in the $\mathbf{E}$-frame. 
We can express this changing orientation in terms of changes in angles $\theta$ and $\phi$. The change in colatitude is
\begin{equation}
\dot{\theta} = 
-\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1) - \vert \mitbf{\Psi} \vert \sin\beta \sin\gamma.
\label{eq:theta_competition}
\end{equation}
where $\mitbf{\Psi}$ is expressed in terms of its amplitude $|\mitbf{\Psi}|$ and its orientation angles $\beta$ and $\gamma$.
Equation~\eqref{eq:theta_competition} captures the essential competition between growth and decay of the mismatch angle.
The first term acts to reduce $\theta$ via TPW, while the second term can increase (or decrease) it arbitrarily via
relative motion between the geographic frame and the principal axes of the $\mathbf{E}$-frame.
Comparison with Equation~\eqref{eq:simple_milankovitch} shows that, when $\vert \mitbf{\Psi} \vert$ is zero,
then the rate of TPW $\vert \dot{\Theta} \vert$ is identical to $\vert \dot{\theta} \vert$.
Conversely, when the rate of TPW is very slow (i.e. $\vert \dot{\Theta} \vert \sim 0$), 
then $\vert \dot{\theta} \vert \sim \vert \mitbf{\Psi} \vert$, so the evolution of $\theta$ is driven mostly by $\mitbf{\Psi}$. 
For the purposes of scaling we will omit the orientation factor $\sin \beta \sin \gamma$ and focus on $|\mitbf{\Psi}|$.
In order for the TPW rate to be large, the growth of $\theta$ via $\vert \mitbf{\Psi} \vert$ must, 
at least occasionally, be larger than its decay.

@@

\begin{figure*}
\centering
\includegraphics[width=0.9\textwidth]{figures/perturb.pdf}
\caption{Graphical demonstration of the $\sin{2 \theta}$ theorem of \citet{davis1970rotation}.  Two spheroidal bodies with eigenvalues $\lambda_2 > \lambda_1$ start out with the rotation axis $\Omega$ aligned with the $\lambda_2$ axis. However, on the left the eigengap $\lVert \lambda_2 - \lambda_1 \rVert$ is large, while on the right it is small.  A negative mass perturbation is instantaneously added to both bodies, which effects a small rotation of the principal axes on the left, but a large one on the right.}
\label{fig:perturb}
\end{figure*}

@@

The preceding results clarify the complex relationship between the Rayleigh number, $m$, and the rates of TPW for a convecting planet.
As mantle convection redistributes mass in the planet's interior, the rotational bulge moves around to stay aligned with the principal axes of the convective moment of inertia. 
There is a constant competition between growth of the mismatch angle $\theta$ through 
Equation~\eqref{eq:grow_perturbation} and its relaxation through Equation~\eqref{eq:simple_milankovitch}.
We may plug in the estimate for $\Lambda_{ij}$ (Equation~\eqref{eq:lambda_estimate}) into Equation~\eqref{eq:simple_milankovitch} to find the strikingly simple expression
\begin{equation}
\frac{\kappa}{R^2} \; \dot{\Theta} \sim -\frac{1}{m} \sin{2 \theta}.
\label{eq:simplest_milankovitch}
\end{equation}
Surprisingly, $\Gamma$ and $\mathrm{Ra}$ have completely dropped from the prefactor in the scaling. 
Response timescales for relaxation of the mismatch angle go down at high Rayleigh numbers.
At the same time, however, coherence in the temperature structure goes down, reducing the amount of power in the degree-two part of the field responsible for driving TPW.
That these two effects cancel is something of a coincidence due to the simple estimate of the smallest lengthscales of the problem.
Scalings for lengthscales of convection in fluids with temperature dependent viscosity \citep[e.g.][]{solomatov1995scaling} or pseudoplastic rheology \citep[e.g.][]{korenaga2010scaling} have different functional dependencies on $\mathrm{Ra}$ or additional nondimensional parameters.
However, a common feature in most scalings is that typical lengthscales are still some power-law of Rayleigh number $d \propto \mathrm{Ra}^{-\beta}$.
With this form, our scaling for TPW rate has the following dependence on $\mathrm{Ra}$:
\begin{equation}
\frac{\kappa}{R^2} \; \dot{\Theta} \sim -\frac{\mathrm{Ra}^{1-3\beta}}{m} \sin{2 \theta}.
\label{eq:simplest_milankovitch_defer_scaling}
\end{equation}
In general, $\beta$ is some small number between one-fourth and one-third, so we expect that more complicated estimates for 
$d(\mathrm{Ra})$ will still result in a weak dependence of the prefactor in Equation~\eqref{eq:simplest_milankovitch_defer_scaling} 
on the Rayleigh number.

We then suggest that the most important parameters are $m$, which acts as the brakes on the system, and the mismatch angle $\theta$.
Whereas the prefactor $\mathrm{Ra}^{1-3\beta}$ in Equation~\eqref{eq:simplest_milankovitch} has a weak dependence on $\mathrm{Ra}$, 
the misalignment angle $\theta$ is expected to be strongly dependent on it.

>>>>

\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
  \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
  \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
  \renewenvironment{#1}%
    {\linenomath\csname old#1\endcsname}%
    {\csname oldend#1\endcsname\endlinenomath}}% 
  \newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
    \patchAmsMathEnvironmentForLineno{#1}%
    \patchAmsMathEnvironmentForLineno{#1*}}%
  \AtBeginDocument{%
  \patchBothAmsMathEnvironmentsForLineno{equation}%
  \patchBothAmsMathEnvironmentsForLineno{align}%
  \patchBothAmsMathEnvironmentsForLineno{flalign}%
  \patchBothAmsMathEnvironmentsForLineno{alignat}%
  \patchBothAmsMathEnvironmentsForLineno{gather}%
  \patchBothAmsMathEnvironmentsForLineno{multline}%
  }

@@

\begin{figure*}
\centering
\includegraphics[width=0.9\textwidth]{figures/reference_frames.pdf}
\caption{Relevant vectors and angles for the TPW analysis. The axes $\mathbf{e}_1$, $\mathbf{e}_2$, and $\mathbf{e}_3$ represent the principal axes of the convective moment of inertia $\mathbf{E}$, with associated eigenvalues $\lambda_1$, $\lambda_2$, and $\lambda_3$, respectively. Since the choice of geographic axes is arbitrary, we assume that at this instant the $\mathbf{E}$-frame and the geographic frames are colocated (though at future times they will not be). The angle $\theta$ represents the mismatch between the rotation axis $\mitbf{\omega}$ and the $\mathbf{e}_3$-axis. For illustration, we assume the longitude $\phi$ of the rotation axis is zero so $\mitbf{\omega}$ lies in the $\mathbf{e_1}$-$\mathbf{e_3}$ plane. True polar wander moves the rotation axis towards the $\mathbf{e}_3$-axis. We denote this motion by $\dot{\mitbf{\omega}}$ in the geographic frame. However, time-dependent convection in the mantle may cause a relative rotation between the geographic frame and the $\mathbf{E}$-frame. This relative rotation can be represented as a rotation around the axis $\mitbf{\Psi}$, which is defined by colatitude $\beta$ and longitude $\gamma$. The rotation around the $\mitbf{\Psi}$ axis contributes to the motion of $\mitbf{\omega}$ as seen in the $\mathbf{E}$-frame.  Further discussion of the reference frames and the relative rotation vector $\mitbf{\Psi}$ can be found in Sections~\ref{sec:reference_frames} and \ref{sec:theta}}
\label{fig:reference_frames}
\end{figure*}

@@

It is useful to define a nondimensional eigenvalue difference (or ``eigengap'') $\Lambda_{ij} \equiv (\lambda_i - \lambda_j)/I_0$.  
This quantity represents the size of fluctuations in the convective moment of inertia compared to the reference value.
The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is 
proportional to the ratio of rotational to gravitational forces \citep{munk1960rotation}:
\begin{equation}
(C-A) \sim I_0 m.
\end{equation}

@@

Convection drives a mismatch $\theta$ between the current rotation axis and the principal axis of the convective moment. 
This parameter is crucial for the rate of TPW, as previously noted in Equation~\eqref{eq:scaled_rotation}.  
Much of the debate around the existence and magnitude of TPW on Earth comes down to the question of how big $\theta$ can be \citep{kirschvink1997evidence, steinberger1997changes}.

Two processes control the evolution of $\theta$. Its growth occurs through perturbations in the convective moment and 
its decay occurs by relaxation of the pole towards the maximum moment of inertia.
We can explore these two effects by converting Equation~\eqref{eq:liouville_ode}
from the body-fixed frame to the $\mathbf{E}$-frame (recall that $\mitbf{\Omega}$ defines
the rotation of the body-fixed frame relative to inertial space in Equation~\eqref{eq:euler}).
The $\mathbf{E}$-frame rotates slowly with respect to the geographic frame, 
which we describe by the rotation vector $\mathbf{\Psi}$. 
(To be explicit in our discussion we let $\mitbf{\Psi}$ define the rotation of the $\mathbf{E}$-frame in the geographic frame, 
although this choice is arbitrary as long as we are consistent). The time derivatives of $\mitbf{\omega}$ in the two frames are related by 
\begin{equation}
\dot{\mitbf{\omega}} = \dot{\mitbf{\omega}}^\prime + \mathbf{\Psi}\times \mitbf{\omega}^\prime.
\end{equation}
where primes are used to define quantities in the $\mathbf{E}$-frame. Rearranging gives 
\begin{equation}
\dot{\mitbf{\omega}}^{\prime} = \dot{\mitbf{\omega}} - \mathbf{\Psi}\times \mitbf{\omega}^\prime.
\end{equation}
We now substitute for $\dot{\mitbf{\omega}}$ from Equation~\eqref{eq:liouville_ode}, 
making the additional assumption that the geographic and $\mathbf{E}$-frames are momentarily aligned. 
This assumption is not a serious restriction because our goal is simply to relate the time derivatives of $\mitbf{\omega}$ in the two frames, 
rather than track the evolution of $\mitbf{\omega}$ over time. The expression for $\dot{\mitbf{\omega}}$ becomes
\begin{equation}
 \dot{\mitbf{\omega}}^\prime  = \frac{1}{(C-A)T_1} \left[ \mathbf{E}^\prime \cdot \mitbf{\omega}^\prime - \left( \mitbf{\omega}^\prime \cdot \mathbf{E}^\prime \cdot \mitbf{\omega}^\prime  \right) \mitbf{\omega}^\prime \right] - \mathbf{\Psi}\times \mitbf{\omega}^\prime.
\label{eq:liouville_ode_new_frame}
\end{equation}
The advantage of expressing the rate of TPW in the $\mathbf{E}$-frame is that the entire equation
may be written in terms of the principal moments $\lambda_i$ and the angles from the principal axes.
As in Equation~\eqref{eq:simple_milankovitch}, we make the simplifying assumption that 
$\mitbf{\omega}^{\prime}$ lies in $\mathbf{e}_1^{\prime}-\mathbf{e}_3^{\prime} $ plane (as shown in Figure~\ref{fig:reference_frames}). 
In this case the orientation of $\mitbf{\omega}^\prime$ is defined by colatitude $\theta$ and longitude $\phi=0$. 
The colatitude $\theta$ is precisely the angular misalignment that specifies the rate of TPW in Equation~\eqref{eq:simple_milankovitch}.
We furthermore specify the orientation of $\Psi$
using colatitude $\beta$ and longitude $\gamma$ (see Figure~\ref{fig:reference_frames}).

The time derivative of the unit vector $\mitbf{\omega}^{\prime}$ defines the changing orientation of the rotation axis in the $\mathbf{E}$-frame. 
We can express this changing orientation in terms of changes in angles $\theta$ and $\phi$. The change in colatitude is
\begin{equation}
\dot{\theta} = 
-\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1) - \vert \mitbf{\Psi} \vert \sin\beta \sin\gamma.
\label{eq:theta_competition}
\end{equation}
where $\mitbf{\Psi}$ is expressed in terms of its amplitude $|\mitbf{\Psi}|$ and its orientation angles $\beta$ and $\gamma$.
Equation~\eqref{eq:theta_competition} captures the essential competition between growth and decay of the mismatch angle.
The first term acts to reduce $\theta$ via TPW, while the second term can increase (or decrease) it via
relative motion between the geographic frame and the principal axes of the $\mathbf{E}$-frame.
Comparison with Equation~\eqref{eq:simple_milankovitch} shows that, when $\vert \mitbf{\Psi} \vert$ is zero,
then the rate of TPW $\vert \dot{\Theta} \vert$ is identical to $\vert \dot{\theta} \vert$.
Conversely, when the rate of TPW is very slow (i.e. $\vert \dot{\Theta} \vert \sim 0$), 
then $\vert \dot{\theta} \vert \sim \vert \mitbf{\Psi} \vert$, so the evolution of $\theta$ is driven mostly by $\mitbf{\Psi}$. 
For the purposes of scaling we will omit the orientation factor $\sin \beta \sin \gamma$ and focus on $|\mitbf{\Psi}|$.
In order for the TPW rate to be large, the growth of $\theta$ via $\vert \mitbf{\Psi} \vert$ must, 
at least occasionally, be larger than its decay.

@@

\begin{figure*}
\centering
\includegraphics[width=0.9\textwidth]{figures/perturb.pdf}
\caption{Graphical demonstration of the $\sin{2 \xi}$ theorem of \citet{davis1970rotation}.  Two spheroidal bodies with eigenvalues $\lambda_2 > \lambda_1$ start out with the rotation axis $\Omega$ aligned with the $\lambda_2$ axis. However, on the left the eigengap $\lVert \lambda_2 - \lambda_1 \rVert$ is large, while on the right it is small.  A negative mass perturbation is instantaneously added to both bodies, which effects a small rotation of the principal axes on the left, but a large one on the right.}
\label{fig:perturb}
\end{figure*}

@@

The preceding results clarify the complex relationship between the Rayleigh number, $m$, and the rates of TPW for a convecting planet.
As mantle convection redistributes mass in the planet's interior, the rotational bulge moves around to stay aligned with the principal axes of the convective moment of inertia. 
There is a constant competition between growth of the mismatch angle $\theta$ through 
Equation~\eqref{eq:grow_perturbation} and its relaxation through TPW.
We may plug in the estimate for $\Lambda_{ij}$ (Equation~\eqref{eq:lambda_estimate}) into Equation~\eqref{eq:simple_milankovitch} to find the strikingly simple expression
\begin{equation}
\frac{\kappa}{R^2} \; \dot{\Theta} \sim -\frac{1}{m} \sin{2 \theta}.
\label{eq:simplest_milankovitch}
\end{equation}
Surprisingly, $\Gamma$ and $\mathrm{Ra}$ have completely dropped from the prefactor in the scaling. 
Response timescales for relaxation of the mismatch angle go down at high Rayleigh numbers.
At the same time, however, coherence in the temperature structure goes down, reducing the amount of power in the degree-two part of the field responsible for driving TPW.
That these two effects cancel is something of a coincidence due to the simple estimate of the smallest lengthscales of the problem.
Scalings for lengthscales of convection in fluids with temperature dependent viscosity \citep[e.g.][]{solomatov1995scaling} or pseudoplastic rheology \citep[e.g.][]{korenaga2010scaling} have different functional dependencies on $\mathrm{Ra}$ or additional nondimensional parameters.
However, a common feature in most scalings is that typical lengthscales are still some power-law of Rayleigh number $d \propto \mathrm{Ra}^{-\beta}$.
With this form, our scaling for TPW rate has the following dependence on $\mathrm{Ra}$:
\begin{equation}
\frac{\kappa}{R^2} \; \dot{\Theta} \sim -\frac{\mathrm{Ra}^{1-3\beta}}{m} \sin{2 \theta}.
\label{eq:simplest_milankovitch_defer_scaling}
\end{equation}
In general, $\beta$ is some small number between one-fourth and one-third, so we expect that more complicated estimates for 
$d(\mathrm{Ra})$ will still result in a weak dependence of the prefactor in Equation~\eqref{eq:simplest_milankovitch_defer_scaling} 
on the Rayleigh number.

We then suggest that the most important parameters are $m$, which acts as the brakes on the system, and the mismatch angle $\theta$.
Whereas the prefactor $\mathrm{Ra}^{1-3\beta}$ in Equation~\eqref{eq:simplest_milankovitch_defer_scaling} has a weak dependence on $\mathrm{Ra}$, 
the misalignment angle $\theta$ is expected to be strongly dependent on it.
